\chapter{Kinematics}

We have our position vector

\begin{equation}
	\vec r(t) = (x(t), y(t))
\end{equation}

We use $\vec r$ because it seems natural, it is the direction we are pointing in.

\begin{remark}
	Sometimes when reference to radial $\vec r$ is misleading, we use $\vec x(t)$.
\end{remark}

The change of the vector in space across time sweeps over some \textbf{trajectory}.

\section{Displacement}

\begin{definition}[Displacement]
	The \textit{displacement vector} $\Delta\vec r$ is a measure of where the particle went (which depends on the origin!).

	\begin{equation}
		\Delta\vec r \equiv \vec r_f - \vec r_i = \vec r(t_f) - \vec r(t_i)
	\end{equation}
\end{definition}

\begin{enumerate}
	\item $\norm{\Delta \vec r} \neq$ distance travelled in general
	\begin{itemize}
		\item distance traveled = arc length of trajectory
	\end{itemize}

	\item $\Delta \vec r$ is coordinate independent.
	
	Take two coordinate systems $S$ and $S'$. Let them be defined with the relation $\vec r = \vec r' + \vec R$ where $\vec r$ and $\vec r'$ are vectors in the respective coordinate systems.

	\begin{equation}
		\begin{cases}
			S: &\Delta \vec r = \vec r_f - \vec r_i\\
			S':&\Delta \vec r' = \vec {r_f}' - \vec {r_i}'
		\end{cases}
	\end{equation}

	If we plug in the relation, we realize that they are the same, $\Delta \vec r = \Delta \vec r'$
\end{enumerate}

\section{Velocity}

\begin{definition}[Average Velocity]
	\begin{equation}
		\vec v_{\mathrm{avg}} \equiv \frac{\Delta \vec r}{\Delta t}
	\end{equation}
\end{definition}

Let $\D \vec r$ be the infinitesimal displacement.

When we consider a smaller interval:

\begin{equation}
	\lim_{\Delta t \to 0} \implies \norm{\D \vec r} = \D r \quad \text{(distance traveled)}
\end{equation}

A small change to $t$ results in a small change in $\D S$ (the distance / speed), proportionally

\begin{align}
	&\D S \propto \D t\\
	\implies &\D S = \left(\frac{\D S}{\D t}\right) \D t
\end{align}

\begin{definition}[Velocity]
	AKA the \textit{instantaneous velocity}

	\begin{equation}
		\vec v (t) \equiv \frac{\D \vec r}{\D t}
	\end{equation}

	\begin{itemize}
		\item $\norm{\vec v}$ = speed
		\item $\hat v$ = direction of motion
	\end{itemize}
\end{definition}

\begin{remark}
	A note on average velocity:

	\begin{equation}
		\vec v_{\mathrm{avg}} = \frac{1}{\Delta t} \int_{t_i}^{t_f} \vec v(t) \D t = \frac{1}{\Delta t} \int_{t_i}^{t_f} \frac{\D \vec r}{\D t} \D t = \frac{\Delta \vec r}{\Delta t}
	\end{equation}

	Note also if we find the magnitude, it would not be the same as the average speed since the norm would go over the integrals instead of what is being integrated.
\end{remark}

\begin{itemize}
	\item $\vec v$ a vector, so write $\vec v(t) = \dot x \hat x + \dot y \hat y = \dot{\vec{r}}$
	\item Compare to frames of reference, $S$ \& $S'$
	
	Suppose $\dot{\vec{R}} \neq 0$.

	Then we have

	\begin{equation}
		\begin{cases}
			\vec r &= \vec r' + \vec R\\
			\vec v &= \vec v' + \vec V
		\end{cases}
	\end{equation}

	This is known as the Galilean transformations, which, at higher velocities, ``translates'' to the Lorentz transformations.
\end{itemize}

We can also obtain $\vec r(t)$ given $\vec v(t)$

\begin{equation}
	\Delta \vec r = \int \D \vec r = \int_{t_i}^{t_f} \vec v \D t
\end{equation}

and

\begin{align}
	\vec r(t) &= \vec r_i + \vec v_i (t - t_i)
\end{align}

\section{Acceleration}

\begin{definition}
	\begin{equation}
		\vec a(t) = \frac{\D \vec v}{\D t} = \frac{\D^2 \vec r}{\D^2 t}
	\end{equation}
\end{definition}

Similar to what is mentioned in \cref*{sect:calc-with-vec}, $\vec a_\parallel$ is change in speed, $\vec a_\perp$ is change in direction of motion.

\begin{remark}
	We do have the \textit{jerk}, but it just seems that it never really matters, and acceleration is fully sufficient.
\end{remark}

\subsection{Cartesian Coordinates}

\begin{equation}
	\begin{cases}
		\vec r(t) &= x(t)\hat x + y(t)\hat y + z(t)\hat z\\
		\vec v(t) &= \dot x(t)\hat x + \dot y(t)\hat y + \dot z(t)\hat z\\
		\vec a(t) &= \ddot x(t)\hat x + \ddot y(t)\hat y + \ddot z(t)\hat z
	\end{cases}
\end{equation}

\begin{example}
	Suppose particle's position is $\vec r(t) = A(e^{\alpha t} \hat x + e^{-\alpha t} \hat y)$ with $A$ and $\alpha$ constants. ($[A] = \mathrm{m}, [\alpha] = \mathrm{m}^{-1}$) Find $\vec v(t)$ and $\vec a(t)$ and sketch trajectory.
\end{example}

\begin{sol}
	\textbf{Velocity}

	\begin{align}
		\vec v(t) &= \frac{\D \vec r}{\D t}\\
		&= A(\alpha e^{\alpha t} \hat x - \alpha e^{-\alpha t} \hat y)\\
		&= \alpha A(e^{\alpha t} \hat x - e^{-\alpha t} \hat y)
	\end{align}

	\textbf{Acceleration}

	\begin{align}
		\vec a(t) &= \frac{\D \vec v}{\D t}\\
		&= \alpha^2 A(e^{\alpha t} \hat x + e^{-\alpha t} \hat y)\\
		&= \alpha^2 \vec r(t)
	\end{align}

	\textbf{Speed}

	\begin{align}
		\abs{\vec v} &= \sqrt{\vec v \cdot \vec v}\\
		&= \sqrt{(\alpha A)^2 \left[e^{2\alpha t} + e^{-2\alpha t}\right]}\\
		&= \alpha A \sqrt{2\cosh(2\alpha t)}
	\end{align}

	Note that (by definition)

	\begin{equation}
		\begin{cases}
			x(t) &= Ae^{\alpha t}\\
			y(t) &= Ae^{-\alpha t}
		\end{cases}
	\end{equation}

	We can try to find $y(x)$ by eliminating $t$, which is the equation for the trajectory, we obtain:

	\begin{equation}
		y(x) = \frac{A^2}{x} \qquad y \propto \frac{1}{x}
	\end{equation}

	So although the velocity and acceleration changes at a exponential rate, the trajectory that it produces exhibits the inverse curve.
\end{sol}

\begin{example}
	A particle moves in the plane with trajectory of a circle of radius $R$. The particle sweeps out the circle at a uniform and constant rate. That is, it undergoes uniform circular motion. Find $\vec r(t), \vec v(t),$ and $\vec a(t)$.
\end{example}

\begin{sol}
	% diagram

	We know that the magnitude of the position vector $\abs{\vec r} = R$ and that

	\begin{equation}
		\vec r = R\cos\theta(t) \hat x + R\sin\theta(t) \hat y
	\end{equation}

	\begin{remark}
		$\vec v$ changes direction, but with uniform rate $\abs{\vec v} = c$.
	\end{remark}

	From our $\vec r(t)$ we have that

	\begin{align}
		\vec v(t) &= -R\sin\theta(t) \left(\frac{\D \theta}{\D t}\right) \hat x + R\cos\theta(t) \left(\frac{\D \theta}{\D t}\right) \hat y\\
		&= R\dot\theta \left[-\sin\theta \hat x + \cos\theta \hat y\right]
	\end{align}

	We know that $v$ is constant and that $v = R\dot\theta$, so $R\dot\theta$ must also be constant. Since $R$ is constant, $\dot\theta$ is constant.

	\begin{equation}
		\dot\theta \equiv \omega \implies \theta(t) = \omega t
	\end{equation}

	This is assuming $\theta(0) = 0$.

	As a result of our derivation, we find

	\begin{equation}
		\begin{cases}
			\vec r(t) &= R\cos(\omega t)\hat x + R\sin(\omega t) \hat y\\
			\vec v(t) &= -\omega R\sin(\omega t) \hat x + \omega R\cos(\omega t) \hat y
		\end{cases}
	\end{equation}

	Now, noting the magnitude:

	\begin{equation}
		\begin{cases}
			r &= R\\
			v &= \omega R\\
			a &= \omega^2 R = \frac{v^2}{R}
		\end{cases}
	\end{equation}

	\textbf{Acceleration}

	\begin{align}
		\vec a(t) &= -\omega^2 R \cos(\omega t) \hat x - \omega^2 R \sin(\omega t) \hat y\\
		&= -\omega^2 \vec r(t)
	\end{align}

	\begin{remark}
		Because $\hat a$ points towards the origin [$\hat a = - \hat r$], we call it ``centripetal'' ($\leftarrow$ central seeking).
	\end{remark}
\end{sol}

\section{Formal Solution of Kinematic Equations}

We want to obtain $\vec v(t)$ and $\vec r(t)$ given $\vec a(t)$.

\subsection{$\vec v$ from $\vec a$}

\begin{align}
	\int_0^t \vec a(t) \D t' &= \int_{\vec v_0}^{\vec v} \frac{\D \vec v}{\D t'} \D t'\\
	&= \vec v(t) - \vec v_0\\
	\vec v(t) &= \boxed{\vec v_0 + \int_0^t \vec a(t') \D t'}
\end{align}

\subsection{$\vec r$ from $\vec v$ (from $\vec a$)}

\begin{align}
	\int_0^t \vec v(t') \D t' &= \int_{\vec r_0}^{\vec r} \frac{\D \vec r}{\D t} \D t\\
	&= \vec r(t) - \vec r_0\\
	\vec r(t) &= \boxed{\vec r_0 + \int_0^t \vec v(t') \D t'}\\
	&= \vec r_0 + \int_0^t \left[\vec v_0 + \int_0^{t'} \vec a(t'') \D t''\right] \D t'\\
	\vec r(t) &= \boxed{\vec r_0 + \vec v_0 t + \int_0^t \int_0^{t'} \vec a(t'') \D t'' \D t'}
\end{align}

\begin{remark}
	We need to know $\vec r_0$.

	To find $\vec r(t)$ given $\vec a(t)$ we need also to know the initial conditions, $\vec r_0$ and $\vec v_0$.
\end{remark}

\section{Constant Acceleration Motion}

There are many cases of constant $\vec a$ motion. With our previous analysis, the cases of when $\vec a =$ const gives:

\begin{equation}
	\begin{cases}
		\vec v(t) &= \vec v_0 + \int_0^t \vec a \D t' = \vec v_0 + \vec a t\\
		\vec r(t) &= \vec r_0 + \vec v_0 t + \frac{1}{2} \vec a t^2\\
		v^2 &= {v_0}^2 + 2 \vec a \cdot \Delta \vec r
	\end{cases}
\end{equation}

\begin{remark}
	if $t_0 \neq 0$, the $t \to \Delta t$ in formulas.
\end{remark}

Let's eliminate $t$ from these equations:

From $\vec v = \vec v_0 + \vec a t$, compute $v^2 = \vec v \cdot \vec v$

\begin{align}
	v^2 &= {v_0}^2 + 2 \vec v_0 \cdot \vec a t + a^2t^2\\
	\frac{1}{2} v^2 &= \frac{1}{2} {v_0}^2 + \vec v_0 \cdot \vec a t + \frac{1}{2}a^2t^2
\end{align}

Now, from $\vec r$ compute

\begin{equation}
	\vec a \cdot \vec r = \vec a \cdot \vec r_0 + \vec a \cdot \vec v_0 t + \frac{1}{2} a^2 t^2
\end{equation}

Then, we take the difference, we have

\begin{align}
	\frac{1}{2} v^2 - \vec a \cdot \vec r &= \frac{1}{2} {v_0}^2 - \vec a \cdot \vec r_0\\
	\frac{1}{2} v^2 &= \frac{1}{2} {v_0}^2 + \vec a \cdot (\vec r - \vec r_0)\\
	\Aboxed{v^2 &= {v_0}^2 + 2 \vec a \cdot \Delta \vec r}
\end{align}

\subsection{Components of the Equations}


