\chapter{Foreword: A Note \& A Disclaimer}
% \addcontentsline{toc}{chapter}{Foreword: A Little Something \& A Disclaimer}

\section{About This Guide} 

This guide is a barebone overview of the equations and concepts in \textit{AP Physics C: E\&M}, and, hopefully, one that is accurate. I made this guide while reviewing for the exam myself...

Feynman -- great physicist, very good at explaining things. There came to be something called the Feynman technique, which is to explain things clearly for the sake of better/solid understanding of material. I wish that by writing this set of notes, I can defog any of my understanding still in the clouds. If I can convince myself that the ``system'' made sense, I'm sure that is more than enough.

Happy Physics-ing!

\begin{flushright}
    Martin Gong\\
    April 10, 2022
\end{flushright}

\section{A Little Disclaimer}

As a student (currently) only attending high school physics, I understand the limitations I face when writing out explanations for myself -- there are a lot of concepts in Physics that I am not aware of. These absences of knowledge can easily lead to incorrect interpretations of concepts not fully discussed during my course of study. Given that this is the first time that I have come in touch with E\&M, I might wound up with some inaccurate descriptions of concepts, which I do not have the ability to rectify.

Any changes/errata for any content is appreciated.

\bigskip

\rightline{\href{mailto:jector.io@outlook.com}{jector.io@outlook.com}}